\chapter*{Acknowledgments}

The Xevolver Project is supported by JST CREST Research Area
``Development of System Software Technologies for post-Peta Scale High
Performance Computing'' led by Dr.~Yonezawa at RIKEN AICS and later by
Dr.~Sato at RIKEN AICS.

XevXML is under active development, and many people are being involved
in the design and development.

Shoichi Hirasawa is a research associate employed by Tohoku University
for the Xevolver Project.  He has been actively working on writing
translation rules and testing XevXML tools. That is, he is always the
first user of the tools whenever new versions are committed to the code
repository.  His feedbacks are always helpful to improve the
tools. Moreover, he is the main author of most rules in the
transformation rule library. He designed the template of XSLT rules for
AST transformation.

There are many people I would like to thank for their contributions:
\begin{itemize}
 \item Contributing Collaborators: \\
       Reiji Suda~(The University of Tokyo),\\
       Yasuharu Hayashi~(NEC Corporation),\\
       Ryusuke Egawa~(Tohoku University),\\
       Daisuke Takahashi~(University of Tsukuba), \\
       Fumihiko Ino~(Osaka University), and \\
       Kazuhiko Komatsu~(Tohoku University)

 \item Students: \\
       Chunyan Wang (Tohoku University),\\
       Xiong Xiao (Tohoku University), and \\
       Daichi Sato (Tohoku University)

 \item Advisory Boards:\\
       Hiroaki Kobayashi~(Tohoku University),\\
       Michael M. Resch~(HLRS),\\
       Wenmei W. Hwu~(UIUC), and\\
       Chisachi Kato~(The University of Tokyo)
\end{itemize}

%\noindent%
